How to select predictive models for decision-making or causal inference?

### 2024-06-02 how_to_select.tex
iterr gigascience format

%%%CONTEMPORARY%%%
\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

%\onecolumn % for one column layouts

%\usepackage{showframe}

@@

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

% added commands and packages
\newtheorem{assumption}{Assumption}
\newtheorem{proposition*}{Proposition}
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\eg}{\emph{e.g.}}
\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
\usepackage{mathtools}
\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}
\usepackage{booktabs} % For formal tables (eg. bottomrules command)
\usepackage{threeparttable} % For table 
\usepackage{caption} % for subfigures
\usepackage{subcaption}
\usepackage{placeins} % for FloatBarrier command to prevent a figure to pass some point.
\usepackage{titlesec} % for title formatting
\titleformat*{\subsection}{\color{jnlclr}\bfseries} % force colored subsection font for paragraphs
\titleformat*{\paragraph}{\bfseries} % force bold font for paragraphs

@@

@@

\journaltitle{GigaScience}
\DOI{DOI HERE}
\copyrightyear{2024}
\pubyear{2024}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[How to select predictive models for decision-making or causal inference?]{How to select predictive models for decision-making or causal inference?}

\author[1, 2 $\ast$]{Matthieu Doutreligne}
\author[1]{Gaël Varoquaux}

\authormark{Author Name et al.}

\address[1]{\orgdiv{Soda}, \orgname{Inria, Saclay}, \orgaddress{\country{France}}}
\address[2]{\orgdiv{Mission Data}, \orgname{Haute Autorité de Santé}, \orgaddress{\country{France}}}

\corresp[$\ast$]{Corresponding author. \href{email:matt.dout@gmail.com}{matt.dout@gmail.com}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

%\abstract{
%\textbf{Motivation:} .\\
%\textbf{Results:} .\\
%\textbf{Availability:} .\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
%online.}

\abstract{\textbf{Background:}
    We investigate which procedure selects the predictive model most
    trustworthy to reason on the effect of an intervention and support
    decision making.\\
    %As predictive models --\eg from machine learning-- give likely
    %outcomes, they may be used to, a
    %causal-inference task. The increasing complexity of health data has opened
    %the door to a plethora of models, but also the Pandora box of model
    %selection: which of these models yield the most valid causal estimates? 
    \textbf{Methods:}
    We study such a variety of model selection procedures in practical settings:
    finite samples settings and without theoretical assumption of
    \emph{well-specified} models. Beyond standard cross-validation or
    internal validation procedures, we also study elaborate causal risks.
    These build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. We evaluate whether
    empirically estimated nuisances, which are necessarily noisy,
    add noise to model selection. We compare different metrics for
    causal model selection in an extensive empirical
    study based on a simulation and three healthcare datasets
    based on real covariates. \\
    \textbf{Results:} Among all metrics, the mean squared error,
    classically used to evaluate predictive modes,
    is worse. Re-weighting it with propensity score
    does not bring much improvements. The
    $R\text{-risk}$, which uses as nuisances a model of
    mean outcome and propensity scores, leads to the best performances.
    Nuisance corrections are best estimated with flexible estimators such
    as a super learner.
    \\
    \textbf{Conclusions:} When predictive models are used to
    reason on the effect of an intervention, they must be evaluated
    with different procedures than standard predictive
    settings; using the $R\text{-risk}$ from causal inference.}
\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

\maketitle

@@

Model selection for causal inference becomes more and more difficult with
increasingly different treated and control populations (Figure
\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
coefficient correlation with $\tau\text{-risk}$ drops from 0.9
(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
(\ref{apd:experiments:additional_results}).

\begin{figure}[!h]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
    \end{minipage}
    \caption{\textbf{Model selection is harder for low population
            overlap}:
        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
        Kendall's and continuous overlap values in Figure
            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
\end{figure}

@@

@@

\onecolumn

\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\setcounter{figure}{0}
\setcounter{table}{0}

\begin{appendices}

    %\part*{Supplementary materials:\\How to select predictive models for decision-making or causal inference? }

    \setcounter{secnumdepth}{3}

@@

    \section{Variability of ATE estimation on ACIC
      2016}\label{apd:toy_example:acic_2016_ate_variability}

@@

    \section{Prior work : model selection for outcome modeling (g-computation)}\label{apd:prior_work}

@@

    \section{Causal assumptions}\label{apd:causal_assumptions}

@@

    \section{Definitions of feasible risks}\label{def:feasible_risks}

@@

    \section{Proofs: Links between feasible and oracle risks}\label{apd:proofs}

@@

    \subsection{Reformulation of the $R\text{-risk}$ as reweighted
        $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}

@@

    \subsection{Interesting special cases}\label{apd:theory:special_cases}

@@

    \section{Measuring overlap}\label{apd:motivation_ntv}

@@

    \section{Experiments}

    \subsection{Details on the data generation process}
    \label{apd:experiments:generation}

@@

    \subsection{ Details on the semi-simulated datasets}\label{apd:experiments:datasets}

@@

    \subsection{Model selection procedures}

@@

    \subsection{Additional Results}\label{apd:experiments:additional_results}

    \paragraph{Definition of the Kendall's tau, $\kappa$}

@@

    \section{Data split choices}\label{apd:results:data_split}

    \subsection{Use 90\% of the data to estimate outcome models, 10\% to
        select them}

@@

    \subsection{Heterogeneity in practices for data split}\label{apd:results:k_fold_choices}

@@

\end{appendices}

\section{Competing interests}
No competing interest is declared.

>>>>

%%%CONTEMPORARY%%%
\documentclass[a4paper,num-refs]{oup-contemporary}%

%%% Journal toggle; only specific options recognised.
%%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
\journal{gigascience}

\usepackage{graphicx}
\usepackage{siunitx}

%%% Flushend: You can add this package to automatically balance the final page, but if things go awry (e.g. section contents appearing out-of-order or entire blocks or paragraphs are coloured), remove it!
% \usepackage{flushend}

@@

% added commands and packages
\newtheorem{proof}{Proof}
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
\newtheorem{proposition}[theorem]{Proposition}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{proposition*}{Proposition}
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\eg}{\emph{e.g.}}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm, algorithmicx, algpseudocode}
\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
\usepackage{mathtools}
\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}
\usepackage{booktabs} % For formal tables (eg. bottomrules command)
\usepackage{multirow} % For multirow in tables
\usepackage{threeparttable} % For table 
\usepackage{caption} % for subfigures
\usepackage{subcaption}
\usepackage{placeins} % for FloatBarrier command to prevent a figure to pass some point.
\usepackage{titlesec} % for title formatting
%\titleformat*{\subsection}{\color{jnlclr}\bfseries} % force colored subsection font for paragraphs
%\titleformat*{\paragraph}{\bfseries} % force bold font for paragraphs

@@

\title{How to select predictive models for decision-making or causal inference?}

%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
\author[1,\authfn{1}]{Matthieu Doutreligne}
\author[2]{Gaël Varoquaux}

\affil[1]{Soda, Inria Saclay, France}
\affil[2]{Mission Data, Haute Autorité de Santé, France}

%%% Author Notes
\authnote{\authfn{1}matt.dout@gmail.com; matthieu.doutreligne@inria.fr}

%%% Paper category
\papercat{Paper}

%%% "Short" author for running page header
\runningauthor{Doutreligne et al.}

%%% Should only be set by an editor
\jvolume{00}
\jnumber{0}
\jyear{2024}

@@

\begin{frontmatter}
    \maketitle
    \begin{abstract}
        \textbf{Background:}
        We investigate which procedure selects the predictive model most
        trustworthy to reason on the effect of an intervention and support
        decision making.\\
        %As predictive models --\eg from machine learning-- give likely
        %outcomes, they may be used to, a
        %causal-inference task. The increasing complexity of health data has opened
        %the door to a plethora of models, but also the Pandora box of model
        %selection: which of these models yield the most valid causal estimates? 
        \textbf{Methods:}
        We study such a variety of model selection procedures in practical settings:
        finite samples settings and without theoretical assumption of
        \emph{well-specified} models. Beyond standard cross-validation or
        internal validation procedures, we also study elaborate causal risks.
        These build proxies of the causal error using ``nuisance''
        re-weighting to compute it on the observed data. We evaluate whether
        empirically estimated nuisances, which are necessarily noisy,
        add noise to model selection. We compare different metrics for
        causal model selection in an extensive empirical
        study based on a simulation and three healthcare datasets
        based on real covariates. \\
        \textbf{Results:} Among all metrics, the mean squared error,
        classically used to evaluate predictive modes,
        is worse. Re-weighting it with propensity score
        does not bring much improvements. The
        $R\text{-risk}$, which uses as nuisances a model of
        mean outcome and propensity scores, leads to the best performances.
        Nuisance corrections are best estimated with flexible estimators such
        as a super learner.
        \\
        \textbf{Conclusions:} When predictive models are used to
        reason on the effect of an intervention, they must be evaluated
        with different procedures than standard predictive
        settings; using the $R\text{-risk}$ from causal inference.
    \end{abstract}

    \begin{keywords}
        Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning
    \end{keywords}

\end{frontmatter}

%%% Key points will be printed at top of second page
% \begin{keypoints*}
%     \begin{itemize}
%         \item This is the first point
%         \item This is the second point
%         \item One last point.
%     \end{itemize}
% \end{keypoints*}

@@

Model selection for causal inference becomes more and more difficult with
increasingly different treated and control populations (Figure
\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
coefficient correlation with $\tau\text{-risk}$ drops from 0.9
(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
(\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}).

\begin{figure}[!h]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
    \end{minipage}
    \caption{\textbf{Model selection is harder for low population
            overlap}:
        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. Supplementary materials presents results for all
        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
        Kendall's and continuous overlap values in Figure
            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
\end{figure}

@@

%\renewcommand\thefigure{S\arabic{figure}}
%\renewcommand\thetable{S\arabic{table}}
%\setcounter{figure}{0}
%\setcounter{table}{0}

\section{Availability of source code and requirements}

Lists the following:
\begin{itemize}
    \item Project name: Caussim
    \item Project home page: \url{https://github.com/soda-inria/caussim}
    \item Operating system(s): Platform independent
    \item Programming language: Python
    \item License: BSD 3-Clause License
\end{itemize}

@@

\onecolumn
\setcounter{secnumdepth}{2}

\appendix
\section{Additional files}

%\part*{Supplementary materials:\\How to select predictive models for decision-making or causal inference? }
%\setcounter{secnumdepth}{1}
%\renewcommand{\thesubsection}{\Alph{subsection}}
%\setcounter{secnumdepth}{3}

@@

\subsection{Variability of ATE estimation on ACIC
    2016}\label{apd:toy_example:acic_2016_ate_variability}

@@

\subsection{Prior work : model selection for outcome modeling (g-computation)}\label{apd:prior_work}

@@

\subsection{Causal assumptions}\label{apd:causal_assumptions}

@@

\subsection{Definitions of feasible risks}\label{def:feasible_risks}

@@

\subsection{Proofs: Links between feasible and oracle risks}\label{apd:proofs}

@@

\subsubsection{Reformulation of the $R\text{-risk}$ as reweighted
    $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}

@@

\subsubsection{Interesting special cases}\label{apd:theory:special_cases}

@@

\subsection{Measuring overlap}\label{apd:motivation_ntv}

@@

\subsection{Experiments}

\subsubsection{Details on the data generation process}
\label{apd:experiments:generation}

@@

\subsubsection{ Details on the semi-simulated datasets}\label{apd:experiments:datasets}

@@

\subsubsection{Model selection procedures}

@@

\subsubsection{Additional Results}

\paragraph{Definition of the Kendall's tau, $\kappa$}\label{apd:experiments:additional_results}

@@

\subsection{Data split choices}\label{apd:results:data_split}

\subsubsection{Use 90\% of the data to estimate outcome models, 10\% to
    select them}

@@

\subsubsection{Heterogeneity in practices for data split}\label{apd:results:k_fold_choices}

@@

\section{Declaration}
\subsection{Competing interests}
No competing interest is declared.
